A Methodology and Implementation of Automated Emissions Harmonization for Use in Integrated Assessment Models

### 2018-01-31 paper.tex
address most of rev 2 comments

\begin{keyword}
Integrated Assessment Models, Climate Change, Harmonization, Air Pollution 
\end{keyword}

@@

In previous studies \cite{meinshausen_rcp_2011,rogelj_discrepancies_2011}, two
\textit{families} of methods have been used: those that operate on the ratio of
base year values (i.e., $\frac{h(t_i)}{m(t_i)}$) and those that operate on the
offset of base year values (i.e., $h(t_i) - m(t_i)$). A number of the classic
methods are implemented in \code{aneris} including ratio-convergence shown in
Equation \ref{eqs:ratio}, offset-convergence shown in Equation \ref{eqs:offset},
and interpolation shown in Equation \ref{eqs:interpolate}. The convergence
factor, $\beta$, scales linearly from 1 to 0 over $[t_i, t_f)$ and is shown in
  Equation \ref{eqs:factor}. In all equations the region, species, and sector
  indices have been dropped for clarity. Each equation is a function of time,
  model trajectory, historical trajectory, base year ($t_i$), and a convergence
  year ($t_f$), at which point the harmonized trajectory converges to the
  unharmonized trajectory. \codeb{aneris} provides a number of methods to choose
  from for each of the harmonization families. A summary of all available
  methods is provided in Table \ref{tab:meths}.

@@

@@

\begin{table}[h!]
\centering
\caption{All Harmonization Methods Provided in \code{aneris}}
\label{tab:meths}
\begin{tabular}{|l|l|l|}
\hline
Method Name                             & Harmonization Family & Convergence Year\\
\hline
\code{constant\_ratio}                  & ratio              & $t_f = \infty$\\
\code{reduce\_ratio\_<year>}            & ratio              & $t_f = $\code{<year>}\\
\code{constant\_offset}                 & offset             & $t_f = \infty$\\
\code{reduce\_offset\_<year>}           & offset             & $t_f = $\code{<year>}\\
\code{linear\_interpolate\_<year>}      & interpolation      & $t_f = $\code{<year>}\\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.pdf}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper (purple) branches represent a ``yes''
      response and lower (orange) branches represent a ``no'' response. The
      coefficient of variation, $c_v$, is defined in Equation \ref{eqs:cov},
      $dH$ is defined as $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and
      decision-making thresholds for $c_v$ and $dH$ are described below. Methods
      labeled in \textit{green} are likely to closely match unharmonized
      results, methods in \textit{yellow} will likely somewhat match
      unharmonized results, and methods in \textit{red} can be expected to have
      a large relative difference between harmonized and unharmonized results.}
  \end{center}
\end{figure}

@@

In most cases, however, models do report values in the harmonization
year. Figure \ref{fig:cases} displays a number of example trajectories which
highlight the possible issues resulting from harmonizing model results in
different contexts. When model and historical values are relatively close, a
convergence method is chosen in order to be as representative as possible to the
underlying unharmonized model results (Figure \ref{fig:cases}, Panel
\code{a}). If values are not close, the constant ratio method is chosen in order
to provide reasonable trajectories that still incorporate modeled effects
(Figure \ref{fig:cases}, Panel \code{b}).

Models can additionally report negative emissions in certain contexts which must
be taken into account during harmonization. Such a case is possible for gas
species which can be extracted from the environment and stored, as is the case
for \cotwo in future scenarios with climate mitigation. If a model provides a
trajectory that transitions from positive to negative emissions and base year
results are similar, then a convergence method is used in order to guarantee
capture of this transition in a representative fashion (Figure \ref{fig:cases},
Panel \code{c}). If the discrepancy in base year results is large, it is
possible for a negative trajectory to be inappropriately harmonized to a
positive, but decreasing, trajectory. As such, the constant ratio method is
chosen (Figure \ref{fig:cases}, Panel \code{d}).

@@

Temporal variability of the historical trajectory is also an important
characteristic when considering the choice of harmonization method.  Emissions
from forest and grassland fires, for example, vary from year to year due to a
combination of meteorological conditions and anthropogenic drivers. Land use
emissions in many IAMs are modeled using average emission factors and do not
capture conditions in a specific year. A longer convergence horizon is thus
desired in order to incorporate highly variable historical data with modeled
results as is consistency in harmonization method because the effects are
modeled similarly across regions and species. In order to detect emissions with
a high amount of variation, a measure of the coefficient of variation, $c_v$, of
the first derivative of the historical trajectory is calculated using the
standard deviation, $\sigma$, and the mean, $\mu$, as shown in Equation
\ref{eqs:cov}.

The value of $c_v$ is then tested against a threshold, $\tau_{c_v}$. To
determine this threshold, an analysis of the recent CEDS and LUC historical data
has been performed. Figure \ref{fig:cov} shows the distribution of LUC $c_v$s
and non-LUC $c_v$s as determined for historical data aggregated to the model
regions of 5 different IAMs involved in the SSP process. A threshold value of
$\tau_{c_v} = 20$ has been chosen based on these observations as it
optimally divides the two distributions. Importantly, tails of the LUC and
non-LUC overlap, thus there are both false positives (~7\% of non-LUC
trajectories) and false negatives (~10\% of LUC trajectories). However, as any
regional definition is model dependent and thus any regional aggregation is
possible an automated detection mechanism is necessary.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{cov.pdf}
    \caption[]{
      \label{fig:cov}
      The distribution of $c_v$ values for LUC and non-LUC historical
      trajectories is shown. CEDS historical data \cite{hoesly_historical_2017}
      is used for non-LUC data and \cite{van_marle_historic_2017} is used for
      LUC data. All historical data has been aggregated from countries to IAM
      model regional definitions, and all gas species included in the historical
      data sets are included in the analysis. The solid black line indicates the
      threshold value, $\tau_{c_v}$, used by default in \code{aneris}.  }
  \end{center}
\end{figure}

@@

\section{Results}\label{sec:results}

In order to show a representative cross section of the performance of the
\code{aneris} harmonization procedure, we focus on the harmonization of results
of the IAM MESSAGE-GLOBIOM \cite{fricko_marker_2017}. Two scenarios from the SSP
scenario library \cite{Riahi2017153,Rao2017346} are presented. The SSP2, or
``middle of the road'', scenario (referred to as SSP2-Ref) is chosen to be
analyzed because MESSAGE-GLOBIOM is the marker scenario for this SSP. This SSP2
scenario lies between two RCPs, 6 and 8.5, with a radiative forcing level of
approximately 6.5 Wm$^{-2}$. We additionally present the results for the
SSP2-based mitigation scenario leading to a radiative forcing of 4.5 Wm$^{-2}$
(referred to as SSP2-4.5). The SSP2-45 scenario is chosen because mitigation
technologies and policies are enacted causing a general reduction in pollutants
and GHGs, including (eventual) negative CO2 emissions in some regions and
sectors due to carbon capture and sequestration and afforestation. Figure
\ref{fig:kyoto} shows the different trends of Kyoto Gases in each scenarios.

@@

Figure \ref{fig:co} presents a case in which the magnitude of a trajectory is
distorted. A large discrepancy ($\sim$300\% relative difference) is observed in
the harmonization year for carbon monoxide (CO) emissions in the industrial
sector specifically for the South Asia (SAS) MESSAGE-GLOBIOM region, which
comprises most of the emissions of the Asian subcontinent. The default method
chosen (\code{constant_ratio}) maintains model trends for the region; however,
overall model results are distorted. By applying a \code{constant_offset}
override, the regional trend and magnitude is maintained. With the new
harmonization method for the SAS region, the global trajectory for industrial CO
also is representative the trends seen in the unharmonized trajectory and the
relative importance of the underlying regional trajectories is maintained.

>>>>

\begin{keyword}
Integrated Assessment Models, Harmonization, Greenhouse Gases (GHGs), Air Pollution 
\end{keyword}

@@

In previous studies \cite{meinshausen_rcp_2011,rogelj_discrepancies_2011}, two
\textit{families} of methods have been used: those that operate on the ratio of
base year values (i.e., $\frac{h(t_i)}{m(t_i)}$) and those that operate on the
offset of base year values (i.e., $h(t_i) - m(t_i)$). Both families of functions
depend on a convergence factor, $\beta$, which scales linearly from 1 to 0 over
$[t_i, t_f)$ and is shown in Equation \ref{eqs:factor}. The use of the
  convergence factor implies that the ratio or offset applied in the base year
  ($t_i$) decays to the unharmonized model result (i.e., the convergence factor
  is 0) in the convergence year ($t_f$). In cases where the convergency factor
  is applied over the entire time horizon, the convergence year is taken to be
  $t_f = \infty$.

@@

A number of the classic methods are implemented in \code{aneris} including
ratio-convergence shown in Equation \ref{eqs:ratio}, offset-convergence shown in
Equation \ref{eqs:offset}, and interpolation shown in Equation
\ref{eqs:interpolate}. In all equations the region, species, and sector indices
have been dropped for clarity. Each equation is a function of time, model
trajectory, historical trajectory, base year ($t_i$), and a convergence year
($t_f$), at which point the harmonized trajectory converges to the unharmonized
trajectory. \codeb{aneris} provides a number of methods to choose from for each
of the harmonization families. A summary of all available methods is provided in
Table \ref{tab:meths}.

@@

\begin{table}[h!]
\centering
\caption{All harmonization methods provided in \code{aneris}. A Convergence year of $\infty$ is provided for the \code{constant\_ratio} and \code{constant\_offset} methods are listed as $\beta = 1$ for all model years in both cases.}
\label{tab:meths}
\begin{tabular}{|l|l|l|}
\hline
Method Name                             & Harmonization Family & Convergence Year\\
\hline
\code{constant\_ratio}                  & ratio              & $t_f = \infty$\\
\code{reduce\_ratio\_<year>}            & ratio              & $t_f = $\code{<year>}\\
\code{constant\_offset}                 & offset             & $t_f = \infty$\\
\code{reduce\_offset\_<year>}           & offset             & $t_f = $\code{<year>}\\
\code{linear\_interpolate\_<year>}      & interpolation      & $t_f = $\code{<year>}\\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.pdf}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper (purple) branches represent a ``yes''
      response and lower (orange) branches represent a ``no'' response. The
      coefficient of variation, $c_v$, is defined in Equation \ref{eqs:cov},
      $dH$ is defined as $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and
      decision-making thresholds for $c_v$ and $dH$ are described below.  Where
      present, convergence years of default methods are provided below the
      method name in parentheses.  Methods labeled in \textit{green} are likely
      to closely match unharmonized results, methods in \textit{yellow} will
      likely somewhat match unharmonized results, and methods in \textit{red}
      can be expected to have a large relative difference between harmonized and
      unharmonized results.}
  \end{center}
\end{figure}

@@

In most cases, however, models do report values in the harmonization
year. Figure \ref{fig:cases} displays a number of illustrative example
trajectories which highlight the possible issues resulting from harmonizing
model results in different contexts. Panels \code{a} and \code{b} highlight
examples where model results peak mid-century, behavior that is seen in a number
of scenarios with general emissions mitigation effects, such as pollution
controls applied by developing nations on transport and industry sectors. Panel
\code{a} highlights a case where model base-year values and history are relative
close whereas Panel \code{b} shows a situation where model values and history
are relatively far apart. Panels \code{c} and \code{d} show similar model
trajectories that peak mid-century but also have negative emissions. Models can
report negative emissions in contexts in which a gas species which can be
extracted from the environment and stored, as is the case for \cotwo in future
scenarios with climate mitigation enacted via the deployment of carbon capture
and sequestration technologies or direct carbon capture technologies. Again, the
relative difference with historical values differ between the panels to explore
harmonization method choices in each situation.

When model and historical values are relatively close, a convergence method is
chosen in order to be as representative as possible to the underlying
unharmonized model results (Figure \ref{fig:cases}, Panel \code{a}). If values
are not close, the constant ratio method is chosen in order to provide
reasonable trajectories that still incorporate modeled effects (Figure
\ref{fig:cases}, Panel \code{b}).

If a model provides a trajectory that transitions from positive to negative
emissions and base year results are similar, then a convergence method is used
in order to guarantee capture of this transition in a representative fashion
(Figure \ref{fig:cases}, Panel \code{c}). If the discrepancy in base year
results is large, it is possible for a negative trajectory to be inappropriately
harmonized to a positive, but decreasing, trajectory. As such, the constant
ratio method is chosen (Figure \ref{fig:cases}, Panel \code{d}).

@@

Temporal variability of the historical trajectory is also an important
characteristic when considering the choice of harmonization method.  Emissions
from forest and grassland fires, for example, vary from year to year due to a
combination of meteorological conditions and anthropogenic drivers. Land use
emissions in many IAMs are modeled using average emission factors and do not
capture conditions in a specific year. A longer convergence horizon is thus
desired in order to incorporate highly variable historical data with modeled
results as is consistency in harmonization method because the effects are
modeled similarly across regions and species. In order to detect emissions with
a high amount of variation, a measure of the coefficient of variation, $c_v$, of
the first derivative of the historical trajectory is calculated using the
standard deviation, $\sigma$, and the mean, $\mu$, as shown in Equation
\ref{eqs:cov}. For a single realiziation of $c_v$, the first derivative
information of the entire historical time period is utilized.

@@

The value of $c_v$ is then tested against a threshold, $\tau_{c_v}$. To
determine this threshold, an analysis of the recent CEDS and LUC historical data
has been performed. Figure \ref{fig:cov} shows the distribution of LUC $c_v$s
and non-LUC $c_v$s as determined for historical data aggregated to the model
regions of 5 different IAMs involved in the SSP process: AIM-CGE, IMAGE, GCAM4,
MESSAGE-GLOBIOM, and REMIND-MAGPIE \TODO{add refs}, each of which have varying
definitions of native model regions comprising different collections of
countries. Therefore, each data point comprising Figure \ref{fig:cov} represents
a realization of $c_v$ for a single combination of native model region, sector,
and emissions species \footnote{A full listing of all sectors and species is
  presented in the case-study discussion in Section \ref{sec:results}, Table
  \ref{tab:sp}}. A threshold value of $\tau_{c_v} = 20$ has been chosen based on
these observations as it optimally divides the two distributions. Importantly,
tails of the LUC and non-LUC overlap, thus there are both false positives (~7\%
of non-LUC trajectories) and false negatives (~10\% of LUC
trajectories). However, as any regional definition is model dependent and thus
any regional aggregation is possible an automated detection mechanism is
necessary.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{cov.pdf}
    \caption[]{
      \label{fig:cov}
      The distribution of $c_v$ values for LUC and non-LUC historical
      trajectories is shown. CEDS historical data \cite{hoesly_historical_2017}
      is used for non-LUC data and \cite{van_marle_historic_2017} is used for
      LUC data. All historical data has been aggregated from their native
      spatial resolution (i.e., individual countries) to IAM model regional
      definitions (i.e., collections of countries), and all gas species included
      in the historical data sets are included in the analysis. The solid black
      line indicates the threshold value, $\tau_{c_v}$, used by default in
      \code{aneris}.  }
  \end{center}
\end{figure}

@@

\section{Case Study: Harmonizing Results from a Global IAM}\label{sec:results}

In order to show a representative cross section of the performance of the
\code{aneris} harmonization procedure, we focus on the harmonization of results
of the IAM MESSAGE-GLOBIOM \cite{fricko_marker_2017}. Two scenarios from the SSP
scenario library \cite{Riahi2017153,Rao2017346} are presented. The SSP2, or
``middle of the road'', scenario (referred to as SSP2-Ref) is chosen to be
analyzed because MESSAGE-GLOBIOM is the marker scenario for this SSP. This SSP2
scenario lies between two RCPs, 6 and 8.5, with a radiative forcing level of
approximately 6.5 Wm$^{-2}$. We additionally present the results for the
SSP2-based mitigation scenario leading to a radiative forcing of 4.5 Wm$^{-2}$
(referred to as SSP2-4.5). The SSP2-45 scenario is chosen because mitigation
technologies and policies are enacted causing a general reduction in pollutants
and GHGs, including (eventual) negative \cotwo emissions in some regions and
sectors due to carbon capture and sequestration and afforestation. A scenario in
which negative emissions play a role in mitigation strategies is particularly
important because of the sensitivity of key indicators, such as end-of-century
radiative forcing (which is used to estimate mean global temperature increase),
to the timing and magnitude of net-zero and total negative \cotwo
emissions. Therefore, these two scenarios represent two relative extreme cases
in the use of a harmonization approach and thus provide a case study as to its
general applicability. Figure \ref{fig:kyoto} shows the different trends of
Kyoto Gases in each scenarios.

@@

Figure \ref{fig:co} presents a case in which the magnitude of a trajectory is
distorted. A large discrepancy ($\sim$300\% relative difference) is observed in
the harmonization year for carbon monoxide (CO) emissions in the industrial
sector specifically for the South Asia (SAS) MESSAGE-GLOBIOM region, which
comprises most of the emissions of the Asian subcontinent. The default method
chosen (\code{constant_ratio}) maintains model trends for the region; however,
overall model results are distorted. By applying a \code{constant_offset}
override, the regional trend and magnitude is maintained. With the new
harmonization method for the SAS region, the global trajectory for industrial CO
also is representative of the trends seen in the unharmonized trajectory and the
relative importance of the underlying regional trajectories is maintained.
